An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-10-03 paper.tex
Complete §7.2 (successful document)

>>>>

The trends that we observed in Section~\ref{sec:trends-documents}
highlight that, over time, the complexity of the standardisation process is
increasing: RFCs take longer to produce, and relate to a larger number of
other documents; authors come from a changing and diversifying set of
affiliations and countries; and the volume of email interaction for each
RFC is increasing, with differences in how senior and junior authors and
contributors interact.  We next seek to understand how these factors might
impact the successful deployment of an RFC in-the-wild, and to test if such
success can be predicted.

% IMC 2021 paper Section 4.1

We build on the work of Nikkhah et al.~\cite{nikkhah2017statistical}.
However, while the focus of that paper was on a set of features that were
derived from the standards documents themselves, we augment this with a set
of features derived from the author and email interaction datasets
described in the previous section.

To explore the factors that influence protocol deployment, we follow a
multi-step process:
\begin{enumerate}
  \item We reproduce the logistic regression model in~\cite{nikkhah2017statistical},
    to obtain similar results (as shown in Table~\ref{tbl:mlres}).
    That paper published an expert annotated dataset, labelling RFCs as
    ``successfully deployed'' or not. This pertains to whether the RFC was
    implemented in-the-wild and saw widespread (largely commercial) uptake.
    This dataset covers 251 RFCs published between 1983 and 2011, and
    includes 20 features. We create a baseline logistic regression model
    with $F_1 = 0.762$, $\mathit{AUC} = 0.650$. The results are similar to
    those reported in~\cite{nikkhah2017statistical} ($\mathit{AUC} = 0.670$).

  \item We curate a set of additional \emph{classification features} driven
    by our findings in Section~\ref{sec:trends-documents}. These are derived
    from the RFCs, their authors, and the mailing list interactions. Given
    that these features are only available for RFCs where Datatracker
    metadata is available, we model these features across a subset of the
    original labelled RFCs from~\cite{nikkhah2017statistical}, looking only
    at the 155 RFCs where all features can be calculated. We will show that
    modelling these RFCs with the expanded feature set substantially
    expands the predictive power of the (Step 1) baseline logistic
    regression model, with $F_1 = 0.820$, $\mathit{AUC} = 0.822$.

  \item We then train classification models to predict if an RFC will be
    deployed, using our expanded feature set.  We test the model with the
    manually labelled RFCs and show that we can successfully classify
    RFCs as deployed or not, with a decision tree-based model having
    $F_1 = 0.822$, $\mathit{AUC} = 0.838$. 
\end{enumerate}

\begin{table}
  \centering
  \begin{tabular}{lrrr}
    \toprule
      Model                              & $F_1$ & $\mathit{AUC}$ & $F1_{\mathrm{macro}}$ \\
    \midrule
      Most frequent class                & .757  &      .500      &     .379     \\
      Baseline                           & .758  &      .616      &     .597     \\
      Baseline + FS                      & .762  &      .650      &     .610     \\
    \midrule
      Most frequent class                & .724  &      .500      &     .379     \\
      Baseline                           & .670  &      .559      &     .547     \\
      Baseline + FS                      & .690  &      .620      &     .563     \\
      Logistic regression all feats      & .728  &      .724      &     .666     \\
      Logistic reg. all feats + FS & .820  &      .822      &     .789     \\
      Decision tree all feats + FS       & .822  &      .838      &     .788     \\
    \bottomrule
  \end{tabular}
\caption{
  Classifier scores on the entire dataset (251 RFCs, above) and those with
  our features available (155 RFCs), with or without feature selection (FS).
}
\label{tbl:mlres}
\end{table}

%..................................................................................................
% IMC 2021 paper Section 4.2
\pb{Classification Features:}
We begin with the set of features that were originally derived in
\cite{nikkhah2017statistical}:
IETF Area; Scope (Local, End-to-End (E2E), Bounded (BN), or Unbounded
(UB)); Type (New (N), New with Incumbent (NI), Backward compatible
extension (EB), or Extension (E)); Change to others (CO); Scalability
(SCAL); Security (SCRT); Performance (PERF); Adds value (AV); and Network
effect (NE). We refer the reader to~\cite{nikkhah2017statistical} for a
full explanation of each of these feature.

We further define an additional set of document-based features, derived
from our characterisations in \S\ref{sec:trends-documents}: days from first
draft to RFC publication; number of drafts before RFC publication; number
of outbound citations to RFCs/Internet-Drafts; page count; number of
inbound citations from articles listed in Microsoft Academic, one and two
years after publication; number of inbound citations from other RFCs
published within one and two years after publication; RFC updates or
obsoletes a previous RFC; keywords per page; and topics (where we use
Latent Dirichlet Allocation (LDA) \cite{blei2003latent} to induce 50 topics
on the texts of all existing RFCs, and use the 50-dimensional probability
distribution over topics for a given RFC as the feature vector).

Next, we derive a set of author-based features, from the work described in
\S\ref{sec:trends-demographics}: number of authors on the RFC; if at least
one author of previously published  RFC; if at least one author in North
America, Europe, or Asia; if at least one author from Cisco, Huawei, or
Ericsson; authors with diverse affiliations; authors in more than one
continent;  at least one academic author; and at least one consultant
author.

Finally, we derive a set of features based on email interactions, from
\S\ref{sec:org-dyn:participants}: number of emails mentioning the
Internet-Drafts that precede publication of the RFC; mean number of emails
sent to all RFC authors, for each sender contribution-duration category
(young, mid-age,  senior); mean number of contributors sending emails to
any RFC author, for each contribution-duration category; number of emails
sent to most junior and senior RFC authors, for each sender
contribution-duration category; and number of contributors sending emails
to most junior and senior RFC authors, for each contribution-duration
category.

%..................................................................................................
% IMC 2021 paper Section 4.3
\pb{Modelling Methodology:}
With our expanded feature set, we next build a classifier to predict the
success (i.e., deployment) of RFCs.

Our expanded feature set, including all variants of each feature, is very
large, totalling 177 features. Given that our additional features can only
be calculated for a limited number of data points (155 RFCs),  it would be
infeasible to conduct statistical analysis, and to build a predictive
model, as there are too few data points. To address this, we take a number
of steps to reduce the feature space, while maintaining interpretability:

\begin{enumerate}
  \item Since the largest feature groups are the topics (50) and
    interaction features (54) we reduce both by applying the $\chi^2$ test
    to leave only the top 5 features in each group.

  \item We remove collinearity by using the Variance Inflation Criterion
    (VIF), removing all features with a VIF value above 5.

  \item We apply forward Feature Selection (FS) to identify features of
    high predictive value, following~\cite{nikkhah2017statistical}.
    Starting from an empty feature set, in each iteration of the forward
    procedure, we expand the feature set with the feature that provides the
    largest increase in the $AUC$ score. The procedure ends when there are
    no more unused features that yield score improvements over the current
    feature set. 
\end{enumerate}
The final set of features with and without FS is given in Tables
\ref{tbl:statsnofs} and \ref{tbl:statsfs} respectively.

Using the above feature set, we then train two classification models relying
on logistic regression and a decision tree.  For assessing predictive
performance of the models we use leave-one-out cross-validation, while for
the final statistical analysis we fit a logistic regression model on the
entire dataset and report the statistically significant coefficients (at
significance level $p \leq 0.1$).

\input{tables-prev/imc-2021/lr_fsnone-trim.tex}
\input{tables-prev/imc-2021/lr_fsauc-trim.tex}

\begin{table}
  \centering
  \begin{tabular}{lrrr}
    \toprule
      Model & $F_1$ & $\mathit{AUC}$ & $F1_{macro}$ \\
    \midrule
      Most frequent class                & .757   & .500   & .379 \\
      Baseline                           & .758   & .616   & .597 \\
      Baseline + FS                      & .762   & .650   & .610 \\
    \midrule
      Most frequent class                & .724   & .500   & .379 \\
      Baseline                           & .670   & .559   & .547 \\
      Baseline + FS                      & .690   & .620   & .563 \\
      Logistic regression all feats      & .728   & .724   & .666 \\
      Logistic regression all feats + FS & .820   & .822   & .789 \\
      Decision tree all feats + FS       & .822   & .838   & .788 \\
    \bottomrule
  \end{tabular}
  \caption{
    Classifier scores on the entire dataset (251 RFCs, above) and those
    with our features available (155 RFCs), with or without feature
    selection (FS).
  }
  \label{tbl:mlres}
\end{table}

%..................................................................................................
% IMC 2021 paper Section 4.4
\pb{Evaluative Results:}
Using the trained models, we next evaluate their accuracy and explore which
features are most predictive of RFC deployment.

As evaluative metrics, we use the $F_1$ score and the area under the ROC
curve ($AUC$), as in \cite{nikkhah2017statistical}. We find the standard
$F_1$ score gives overly optimistic performance estimates due the data
being skewed towards the positive class. Consequently, we also report an
$F1_{macro}$ score that takes into account both the positive and negative
class, reflecting performance more realistically. As a baseline comparison,
we present results of our re-implementation of \cite{nikkhah2017statistical}
using their feature set.

The prediction results of the models and feature sets are summarized in
Table \ref{tbl:mlres}. Our baseline model performs at a similar level to
that from \cite{nikkhah2017statistical}.  We further obtain considerable
performance improvements with our additional features. Furthermore, we find
that additional performance can be obtained by using feature selection.
The best performing model, with an F1 score of 0.822, is the decision tree
trained on the entire feature set.
Note, we also tested several non-linear models (neural networks, support
vector machines with non-linear kernels). These attained similar or worse
results as our decision tree model, and therefore we omit them due to space
constraints.

We next investigate the most important features in predicting success.  We
posit that these can offer useful insight for working group chairs and RFC
editors.  The feature importance results are presented in
Tables~\ref{tbl:statsnofs} and \ref{tbl:statsfs}.  Some meaningful patterns
emerge here. We observe features such as adding value, being scalable,
obsoleting other RFCs, page count, increased keyword usage, no incumbent
RFC to compete against and limited scope all being positively correlated
with the likelihood of an RFC becoming deployed. In contrast, we find that
having a broad (unbounded) scope negatively impacts deployment.

We also observe some curious trends that speak to the limitations of our
dataset. The model finds that having an author in Asia is negatively
correlated with deployment, though with only borderline statistical
significance. Further analysis shows that only 10\% of labelled RFCs have
an author in Asia.  As shown in \S\ref{sec:trends-demographics}, we see
that demographics of the IETF are changing, with a recent notable increase
in author representation from Asia. This might suggest that new authors
need time to learn what makes a deployable protocol; equally our deployment
data may be biased towards the Internet in North America and Europe. This
finding requires much more exploration.

We further see that some of the topics extracted by LDA are also useful.
For example, Topic 13 is characterised by a cluster of terms associated
with MPLS, a widely deployed routing protocol. We find a negative
correlation with this topic, but this is likely because there is a
significant number of RFCs that propose modifications or additional
features for the protocol, some of which do not see deployment. Overall,
most of the results are in line with expectations, but more annotated data
is required for further insights.

%..................................................................................................
% IMC 2021 paper Section 4.5
\pb{Discussion:}
% Building on existing work.
Features associated with building on existing RFCs positively correlate
with deployment. RFCs that obsolete earlier versions of the same protocol
are likely to be deployed, indicating that the IETF community is good at
identifying and maintaining protocols that are seeing use. This is also
highlighted by the significance of the inbound RFC citation and ``adds
value'' features. These show that RFCs that are later cited by other
documents, or that add value to other protocols in the stack, are more
likely to be deployed. Protocols that are extensible, and can be extended
and adapted for new uses, are more likely to see deployment. The discussion
in \cite{clark:2002:tussle} on careful choice of extension points
resonates; as do the implications for starting new work.

% Limited scope.
We find that having end-to-end scope (i.e., where only the endpoints of a
connection needs to implement the RFC) is positively correlated with
deployment, whereas having an unbounded scope (i.e., where the entire
Internet may need to be updated) is negatively correlated. This indicates
that well-scoped RFCs, that are cognisant of their deployment challenges,
are more likely to see deployment. A recent IAB workshop \cite{rfc8980}
noted that deployment often occurs in unforeseen ways: limiting the changes
needed to deploy a protocol facilities this.

% Normative requirements.
The number of keywords (e.g., SHOULD, MUST) used per page, which shows the
number of normative requirements an RFC imposes on implementations, also
correlates with deployment. This suggests that well-specified requirements
lead to robust, interoperable, implementations that see wide use.
Unsurprisingly, features such as citation rates are also predictive of
deployment.

% Diversity.
Notwithstanding the limitations of the dataset, we find that the majority
of author demographic features are not significant. While \S\ref{sec:trends-demographics}
identified recent shifts in the demographics of authors, these do not
appear to have a major influence in the deployment of the RFCs that they
write.

% Meeting a need.
Our results broadly support the conclusions of \cite{RFC5218}, where the
IAB noted that, in addition to being technically sound, a successful
protocol must meet a need, be incrementally deployable, and be open to
extension and maintenance.

%..................................................................................................
\pb{Summary:}
Our results show that RFCs that build on existing work, have well-defined
requirements, limited scope, and meet a need have improved chances of being
widely deployed. These document-based features have implications for how
the IETF should design and specify protocols.

We found that the majority of our author-based features, including
geographic and affiliation diversity, did not significantly impact the
deployment prospects of RFCs. Further work is needed to expand the dataset
to ensure that this result holds across a larger set of RFCs.

Finally, the focus of our modelling in this paper has been on the impact of
document, author, and interaction-based features on protocol deployment.
However, this is only one part of the lifecycle of an RFC. It remains to
consider the impact of these, and other, features on the key stages of an
Internet-Draft's development towards becoming an RFC, such as working group
adoption.
